---
title: "LIME example: Costa Rican snapper"
author: "Merrill Rudd"
date: '`r Sys.Date()`'
output: 
  html_document:
    number_sections: yes
    toc: yes
---

# Summary
The `LIME` package was developed to provide an estimate of stock status given the minimum requirements of (a) at least 1 year of length composition and (b) assumptions about biological parameters. The `LIME` package stands out from other length-based assessment methods in its estimation of annual fishing mortality and recruitment as a random effect, explicitly acknowledging process and observation error. Other methods overlook variation in the recruitment and/or fishing mortality processes. Another advantage of `LIME` is the integration of multiple years of length composition data, an abundance index, and/or catch time series, if available. 

This vignette covers the steps in running the LIME method using an example. 

The example in this vignette comes from the Costa Rican spotted snapper fishery. The data provided here is already massaged into the proper format. Please contact Merrill Rudd (merrillrudd@gmail.com) before using or distributing this dataset. 



# Installing and loading the package

Install the R package from github using devtools:
```{r install_packages, message=FALSE}
devtools::install_github("merrillrudd/LIME")
```

Then load the library:
```{r load_packages, message=FALSE}
library(LIME)
```

# Example dataset

An example dataset for the LIME method is the Costa Rican spotted rose snapper. PRETOMA, a Costa Rica-based NGO, has been conducting surveys in of the snapper population in the Gulf of Nicoya for nine years (2007-2015). These surveys took length measurements from the catch, either on board vessels or at the dock. There is also a CPUE index based on the onboard observations of the soak time for the longlines/gillnets, and the number of hooks per longline and mesh size of gillnet, respectively. Since the majority of the catch was from longline gear, we only used the longline information for the CPUE time series. The length composition each year, however, was weighted based on the catch per gear, so as to use all of the length measurement information but not need to estimate two selectivity curves. The surveys, however, do not adequately account for the total catch from this fishery, therefore only the length composition over time, abundance index, and biological information are available for the stock assessment.

```{r explore_data, echo=FALSE}
par(mfrow=c(3,3), mar=c(0,0,0,0), omi=c(1,1,0.7,0.2))
for(i in 1:nrow(snapper$LF)){
  barplot(snapper$LF[i,], xlab="", ylab="", xaxs="i", yaxs="i", col="black", xaxt="n", yaxt="n", xlim=c(0,ncol(snapper$LF)), ylim=c(0,max(snapper$LF)*1.2))
  if(i %in% c(1,4,7)) axis(2, at=pretty(c(0,max(snapper$LF))), las=2, cex.axis=1.2)
  if(i %in% c(7,8,9)) axis(1, at=pretty(c(0,ncol(snapper$LF))), cex.axis=1.2)
  text(x=8, y=(max(snapper$LF)), snapper$years[as.numeric(rownames(snapper$LF)[i])], cex=1.2)
  box()
}
mtext("Length bin (cm)", side=1, line=3, cex=1.2, outer=TRUE)
mtext("Proportion of catch", side=2, line=4, cex=1.2, outer=TRUE)
mtext("Length composition from catch samples", side=3, line=1, cex=1.2, font=2, outer=TRUE)
```

# Running the LIME method
## Compile inputs
### Life history and starting values

Use the function `create_lh_list` to specify life history information for the stock you are assessing. The required inputs include: 

 * *vbk* -- von Bertalanffy growth 
 
 * *linf* -- von Bertlanffy asymptotic length 
 
 * *lwa* -- length-weight scaling parameter 
 
 * *lwb* -- length-weight allometric parameter 
 
 * *S50* -- starting value for age or length at 50% selectivity (will be estimated in model)
 
 * *selex_input* -- specify whether *S50* is "age" or "length" 
 
 * *M50* -- age or length at 50% maturity 
 
 * *maturity_input* -- specify whether *M50* is "age" or "length" 
 

You can also specify the naturality mortality rate (_M_), steepness (_h_), coefficient of variation around the growth curve (_CVlen_), and other values (including observation and process error) if some information is known. Otherwise, default values are specified for these parameters. See the help file `?create_lh_list` for default values of other life history parameters and the relationships used to derive life history information.

```{r create_lh, warning=FALSE}
lh <- create_lh_list(vbk=0.21, linf=64.58, lwa=0.0245, lwb=2.79, S50=30, selex_input="length", M50=34, maturity_input="length", binwidth=1, CVlen=0.1)
```
```{r plot_lh, echo=FALSE}
par(mfrow=c(1,2), mar=c(8,5,1,4))
plot(lh$L_a, xlab="Age", ylab="Length (cm)", xlim=c(1,length(lh$L_a)), ylim=c(0,max(lh$L_a)*1.1), xaxs="i", yaxs="i", type="l", lwd=4, col="black")
par(new=TRUE)
plot(lh$W_a, xlab="", ylab="", xaxs="i", xlim=c(1,length(lh$L_a)), ylim=c(0, max(lh$W_a)*1.1), yaxs="i", xaxt="n", yaxt="n", type="l", lwd=4, lty=2, col="blue")
legend("bottomright", legend=c("Length", "Weight"), col=c("black", "blue"), lwd=4, lty=c(1,2))
axis(4, at=pretty(c(0, max(lh$W_a)*1.1)), col="blue", col.axis="blue")
mtext(side=4, "Weight (g)", col="blue", line=3)

plot(lh$Mat_a, xlab="Age", ylab="Proportion mature", col="forestgreen", type="l", lwd=4, xaxs="i", yaxs="i", ylim=c(0,1.1))
```

### Observed data

In this example, the data is already included in the `snapper` object in the proper format along with the life history information. 

To start from scratch with your own dataset, you will need a tagged list with:

* `years` -- a vector of total years that you want to model. This can be in years (e.g. 2001-2010) or an index of years (1-10). It can also extend backward longer than the data available (e.g. 10 years before the first year of data).
```{r}
input_data <- list()
input_data$years <- snapper$years
input_data$years
```

* `LF` -- a matrix with the observed length frequency (i.e. composition), with the length bins along the columns and the years along the rows. The columns should be named with the upper length bin (e.g. 1-50 for 1 cm length bins). The rows should be named with the index of the year from which the data is collected out of the years for which you want to run the model. In the snapper example, we have data from years 11-19, because we are modeling 10 years prior to the first year of data (1-19 total representing 1987-2005, for which we have data from 2007-2015).
```{r}
input_data$LF <- snapper$LF
input_data$LF[,1:20]
```

If there is only one year of length composition data available, the data should still be in the same matrix format with column and row names in the same structure as with multiple years. Note that the data may need to be manipulated to have length bins along the columns and the year on the rows.
```{r}
## alternate data setup - 1 year of length composition data
LF1 <- t(as.matrix(snapper$LF[nrow(snapper$LF),]))
rownames(LF1) <- rownames(snapper$LF)[nrow(snapper$LF)]
#input_data$LF <- LF1
LF1
```

The effective sample size of observations should also be included. 50 length measurements from the same day, or the same boat, are likely autocorrelated and do not contribute the same amount of information as measurements from 50 days, for example. Using the multinomial likelihood distribution, the effective sample size must be designated. The effective sample size could be estimated in future versions of LIME. In this case, I chose to use the number of days that the fishery was sampled.

```{r}
## specify effective sample size of length measurements
input_data$obs_per_year <- snapper$obs_per_year
input_data$obs_per_year
```


Other input options include: 

* `I_t` -- a vector with abundance index each year, named with the index representing the year from which it came (e.g. index from some years but not all when modeling years 1-19). The following example is the abundance index derived from the snapper dataset. Note that an abundance index is not available in year 15 so that year is skipped in the vector. 
```{r}
input_data$I_t <- snapper$I_t
input_data$I_t
```


* `C_t` -- a vector of catch data each year, named with the index representing the year from which it came (e.g. catch data from years 10-19 when modeling years 1-19).
The following example demonstrates a hypothetical catch of 50 units each year for years 10-19. Note that the catch data does not have to be available every year, and so the year labels would skip the years where the data are missing.

```{r}
#input_data$C_t <- C_t
C_t <- rep(50, length(10:19))
names(C_t) <- 10:19
C_t
```

### Combine to input list

Use the function `create_inputs` to make one list that includes the life history information and observed data inputs. 

The `param` and `val` arguments can be used to make last-minute adjustments to the input values. For example, if you want to try running the model with a steepness `h` fixed below 1, you would specify `param="h"` and `val=0.7` (for example). You can also add a concatenated string of parameter names and values if you want to adjust several values at the same time, for example `param=c("h", "CVlen")` and `val=c(0.7, 0.05)`. The `param` names must match the names of the outputs from `create_lh_list` and `input_data`.

```{r}
inits <- create_inputs(lh=lh, input_data=input_data, param=FALSE, val=FALSE)
names(inits)
```

Your named list `inits` is now ready to plug in assumed values, starting values for parameters to be estimated, and data time series to fit the model. 


## Run model
You will use the `run_LIME` function to run the LIME method. The arguments for applying the LIME method to real data (as opposed to simulation) are:

* `modpath` -- the path to the directory to save results

* `lh` -- the list of life history and starting values created above from `create_lh_list`

* `input_data` -- the tagged list of observed data and years, created above (input_data)

* `est_sigma` -- which variances to estimate? The analyst can specify a concatenated vector of variance parameter names to estimate. Because this is a mixed effects model, the model should at least estimate the recruitment process error, `log_sigma_R`. Another helpful variance to estimate is the CV for the growth curve, called `log_CV_L`. The analyst could also specify `log_sigma_I`, `log_sigma_C`, and/or `log_sigma_F` as the observation errors for the abundance index, catch, or process error on the fishing mortality, respectively. 

* `data_avail` -- types of data included, must at least include LC (length composition). May also include Catch or Index separated by underscore, if an abundance index and/or catch data are available. For example, `LC`, `Catch_LC`, `Index_Catch_LC`. 

* `rewrite` -- if you are running a bunch of models and don't want to re-run the models that already have output in their directory, flag `rewrite=TRUE`. If you want to re-run models that may already have output in the directory, tag `rewrite=FALSE`. This argument will not affect the runs if the model has never been run before. 

* `fix_f` -- year (by index, not actual year) to start estimating fishing mortality (e.g. year 11 out of 20 to get estimates for last 10 years); the value of F in this year will be used as the estimate and SE for all previous years. 0=estimate all years. 

* `simulation` -- set to FALSE if doing an assessment on real data. set to TRUE if using simulated data.
  + `itervec` -- set to NULL if this is not a simulation study. Vector of iterations when using simulated data (e.g. 1:100 will run 100 iterations, 50:100 if adding to 1:50 iterations already run)
  
```{r, eval=FALSE}
run <- run_LIME(modpath=<path>, lh=lh, input_data=input_data, est_sigma=c("log_sigma_R","log_CV_L"), data_avail="LC", rewrite=TRUE, fix_f=0, simulation=FALSE, itervec=NULL)
```


## Check results

Do your results make sense? Check out the tagged list of outputs saved in `Report.rds` and the standard errors in `Sdreport.rds` saved in the results directory. Compare the output from different models (e.g. estimating or fixing `CVlen`).

Files saved in output directory:

* `obsData.rds` -- observed data

* `Inputs1.rds` and `Inputs2.rds` -- inputs into TMB. Check that these look right. The `Inputs2.rds` are the most recent inputs after an initial run that estimated the standard error for the recruitment deviation used to develop the bias adjustment term.

* `Report.rds` -- tagged list of outputs from the population model.

* `Sdreport.rds` -- standard deviations for a select list of estimated and derived parameters from the population model.

* `Derived_quants.rds` -- list of derived quantities such as MSY and SPR-based reference points. 

* `df.csv` -- this shows the final gradient and estimated values for each parameter. Check that the final gradients are less than 0.01. If the model did not converge, there will be a text file called `high_final_gradient.txt` to flag the non-convergence.

## Run sensitivity analyses

Run your analysis with a series of alternate values for the inputs in `create_lh_list` that have uncertainty. For example, you may have assumed natural mortality `M` was approximately 1.5*vbk, the default relationship. Try running the model with a series of possible values for `M`. Create a likelihood profile for each to determine if there is a better fit to the data given the alternate value. 